---
title: "Initial Data Review & Dimorphism"
output: html_notebook
---

Initial Data Review for African Murid Dataset

Current Goals here are:

1. Wrangle and organize downloaded data from original format

2. Filter out species and data that fit our requirements for intraspecific variation measurements


# Data Wrangle
#### Load Packages
```{r}
pacman::p_load(tidyverse, brms, tidybayes, janitor, kableExtra, cmdstanr)
options(brms.backend = "cmdstanr")
scale2 <- function(x, na.rm = TRUE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)
```

#### Load Data
There are two downloaded spreadsheets. The first is the **species** spreadsheet. This includes taxon id, catalog #, nomenclatural info, coordinates, sex, sexual condition, tissue info, etc. The second is the **measurements** spreadsheet. This contains catalog #, taxon, external measurements, sex, and the 25 cranial measurements. 

We need a final spreadsheet that contains both of these combined, and filtered to only include specimens with some measurement data. This means the **measurements** spreadsheet can determine which of the speciomens in the **species** spreadsheet we want to keep. There are some extraneous columns I will filter out when loading.
I am also adding a spreadsheet of locomotor modes. These are 4 categories - Arboreal, General, Terrestrial, and Amphibious - taken from Nations et al 2020. They are only for the Murine rodents in the data
```{r message=FALSE}
#show_col_types = FALSE
spec <- read_csv("data/AfricanMammalDB_SPECIES_JAN_15Dec2021.csv", col_types = cols()) %>% 
  clean_names() %>% 
  select(!c(family_authors, 
            genus_authors, 
            species_authors, 
            subspecies, 
            subspecies_authors,
            determinator,
            determination_year,
            date_collected_end,
            available,
            url,
            trap,
            type,
            accuracy,
            tissues,
            collector,
            altitude_max))

measur <- read_csv("data/AfricanMammalDB_MEASUREMENTS_JAN_15Dec2021.csv", col_types = cols()) %>% 
  clean_names() %>% 
  select(!c(subspecies, url))

loc <- read_csv("data/Loc_Modes.csv") %>% 
  separate(taxon, c('genus', 'species'), sep="_") %>% 
  rename(loc_state = state_name,
         loc_state_numeric = state)
```

#### Combine Data Frames and change to Darwin Traits
We want to combine the two data frames and drop species with no measurements at all. The challenge is that we have two possible matching columns. We first want to match with catalog_number, and then try to match using collector_number. The easiest way to do this is create a new column combining collector number and catalog number, then combine on that, then drop it. 

We also need to remove all non-museum (called "preserved") specimens, and remove specimens that have NA for all measurements. 

```{r}
spec <- spec %>% unite("z", c(collection_number,field_number), remove = FALSE)

dat <- measur %>% 
  unite("z", c(collection_number,field_number), remove = FALSE) %>%
  full_join(spec) %>% 
  select(!z) %>% 
  #remove all non-museum preserved specimens
  filter(basis_of_record == "preserved") %>% 
  #Remove rows with NA for ALL measurements
  filter_at(vars(hb:m25), any_vars(!is.na(.))) %>% 
  # Remove unknown species, and species with "n.sp1" etc
  filter(!is.na(species),
         !is.na(genus),
         !species %in% c("n.sp.1", "n.sp.2")) %>% 
  #Darwin Core Stuff
  add_column(continent = "Africa") %>% 
  rename(massing = weight,
         decimalLatitude = latitude,
         decimalLongitude = longitude,
         headbodylength = hb,
         taillength = tl,
         hindfootlength = hf,
         earlength = el,
         elevation = altitude) %>% 
  # add locomotor mode
  left_join(loc)

rm(spec, measur)
```

#### Save File 
```{r}

dat %>% write_csv("data/AfricanMammalDB_CLEANED_10Jan2022.csv")

```
Some simple summaries to see what we have. I think we want to have at least 20(?) per species to test for intraspecific variation?
```{r}
t1 <- dat %>% group_by(family, genus) %>% count(species) %>% filter(n >= 20) 
t1
t1 %>% kbl() %>% kable_paper()

dat %>% count(genus) 
dat %>% count(family)
dat %>% filter(family == "Muridae") %>% count(species)
```
At the n=20 threshold, we have 122 species from 78 genera and 17 families
Also looks like around 26k data points that have locomotor modes.


Explore the variance within each species and plot
```{r} 
t1 <- dat %>%  
  mutate(s_lnHB = scale2(log(hb)),
         sHB = scale2(hb)) %>% 
  group_by(genus, species) %>%
  summarise(mean_lnHB = mean(s_lnHB, na.rm = TRUE),
            var_lnHB = var(s_lnHB, na.rm = TRUE),
            mean_HB = mean(sHB, na.rm = TRUE),
            var_HB = var(sHB, na.rm = TRUE),
            n = n()) %>% 
  filter(n >= 20,
         !is.na(mean_HB)) %>% 
  arrange(var_lnHB) 
  
t1   %>% kbl() %>% kable_paper()

t1 %>% ggplot(aes(x = mean_lnHB, y = var_lnHB)) + geom_point()

t1 %>% group_by(genus, species) %>% summarise(n = n())
```
This is weird because the variance is strongly correlated with the mean. So.... we need to think about how we are scaling the variables, and how we are comparing the variances. Perhaps each family should be scaled together?

Take a look at the plot with species with var_lnHB > 1 removed
```{r}
t1 %>% filter(var_lnHB <= 1) %>% ggplot(aes(x = mean_lnHB, y = var_lnHB)) + geom_point() + geom_smooth(method = lm)

t1 %>% filter(var_lnHB <= 1) %>% ggplot(aes(x = mean_HB, y = var_HB)) + geom_point() + geom_smooth(method = lm)
```
Still more variance with higher mean values


```{r}
dat %>% group_by(genus, species) %>% 
  summarise(mean = mean(hb, na.rm = TRUE),
            var = var(hb, na.rm = TRUE),
            n = n()) %>%
  filter(#var <= 1.7,
         n <= 1000) %>% 
  ggplot(aes(y = var, x = log(n))) +
  geom_point() +
  ggtitle("Some heteroskedacity but not bad!") +
  theme_bw()
```
# Problems

Here is a list of problems with the dataset as I run across them: 

1. HB Below 10 is pretty much impossible. Really that should extend up to 20 or 30. 
2. HF above 100 is crazy except for a few things like the Hystricomorpha. 
3. Lophuromys zena HF values are all wrong, all are in the 200-270 range. Looks like there is just a missing decimal, so 240 should be 24.0. 
4. There are clearly lots of cropped tails (Dendromus tail = 2??), though it will be hard to know which are cropped, because it's hard to know the range for each species.

Basic modeling play
Only 67 of the species have loco info, so maybe this isn't the best. 
These will take a long time, so I want to try some simple stuff to make sure it's working. Running without the grouping variable is fast. Adding grouping makes it much slower.
```{r}
dat <- unite(taxon, c(genus, species), sep = "_")

m1 <-  brm(bf(headbodylength ~ 0 + loc_state + (1|taxon),
           sigma ~ 0 + loc_state),
          family = student(),
          data = dat,
         # prior = c(prior(student_t(3, 0, 2.5), class = "b"),
                    #prior(normal(0,1), class = "b", dpar = "sigma"),
                    #prior(student_t(3, 0, 2.5), class = "sd")),
          #refresh = 0,
           chains = 4, 
           cores = 4,
          iter = 500)
```

```{r}
m1
```
```{r}
plot(m1, ask=F, N=3)
```



m1 <-  brm(bf(headbodylength ~ 0 + loc_state + (1|species),
             sigma ~ 0 + loc_state),
          family = student(),
          data = dat,
          prior = c(prior(student_t(3, 0, 2.5), class = "b"),
                    prior(normal(0,1), class = "b", dpar = "sigma"),
                    prior(student_t(3, 0, 2.5), class = "sd")),
          refresh = 0,
           chains = 4, 
           cores = 4,
          iter = 5000)