---
title: "Initial Data Review & Dimorphism"
output: html_notebook
---

Initial Data Review for African Murid Dataset

# Sexual Dimorphism and Intraspecific variation

Today I tried to look at differences in body-length between males and females while including taxon as a group-level effect. For some reason the grouping variable messes everything up. Probably due to the sheer number of some of the taxa in the set?

This is left today (3 November 2021) with 2 model versions, one without intraspecific variation, and one with. Interesting they give different answers! Without intra-variation, females are larger. With, males are larger. 

TO-DO:
-) Think about the phylogeny. Is this important? Use the Nations 2020 tree, even if it cuts the # of species by a lot. 
-) Still lousy convergence on the sex means. Is this just due to the range of body sizes? The priors? Not sure. 
-) In addition, why are the means for both males and females above zero??
-) Next I want to attach the locomotor mode info to the traits (the tree will help this!), and start working with intraspecific variance between locomotor modes. I think that arboreal taxa may have a lower intraspecific variance than "terrestrial" taxa. Based on Nations 2020, general were tightly conserved too. This is a great extension to that study. 

 
#### Data Processing
This is pretty easy given the clean state of the data
```{r}
pacman::p_load(tidyverse, brms, tidybayes)
scale2 <- function(x, na.rm = TRUE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)
```

```{r}
dat <- read_csv("data/murid_download_11Nov2021.csv") %>% 
  mutate(taxon =  str_c(genus,"_", species)) %>% 
  drop_na(taxon) %>% 
  ## Remove some problematic HB observatios
  filter(hb < 400 ) %>% filter(hb > 40 ) %>% 
  select(-c("basis of record", subspecies, url)) %>% 
  mutate_if(is.numeric, scale2, na.rm = TRUE)

dat %>% distinct(taxon) %>% count()
```
131 species in the dataframe. Not bad!

```{r}
dat %>% group_by(taxon) %>% 
  drop_na(hb) %>%
  summarise(mean = mean(hb, na.rm = TRUE),
            var = var(hb),
            n = n()) %>% 
  arrange(desc(n))

dat %>% group_by(taxon) %>% 
  drop_na(hb) %>%
  summarise(n = as.numeric(n())) %>% 
  ungroup() %>% 
  mutate( median = median(n),
          mean = mean(n))
```
Good list of variance. Does variance increase or decrease with number of species sampled?
Plot log(n) because there is one taxa, Mastomys_natalensis, with 6269 samples. The next closest is 1459!
```{r} 
dat %>% group_by(taxon) %>% 
  summarise(mean = mean(hb, na.rm = TRUE),
            var = var(hb, na.rm = TRUE),
            n = n()) %>%
  ggplot(aes(y = var, x = log(n))) +
  geom_point()
```
Plot, filtering the taxa with huge variance

```{r}
dat %>% group_by(taxon) %>% 
  summarise(mean = mean(hb, na.rm = TRUE),
            var = var(hb, na.rm = TRUE),
            n = n()) %>%
  filter(var <= 1.7,
         n <= 1000) %>% 
  ggplot(aes(y = var, x = n)) +
  geom_point() +
  ggtitle("Some heteroskedacity but not bad!") +
  theme_bw()
```
Try removing taxa with fewer than 10 samples. This leaves 92 species. 
```{r}
dat %>% group_by(taxon) %>% 
  drop_na(hb) %>%
  summarise(n = n()) %>% 
  filter(n >= 10) %>% 
  nrow()
```
Here's that plot
```{r}
dat %>% group_by(taxon) %>% 
  summarise(mean = mean(hb, na.rm = TRUE),
            var = var(hb, na.rm = TRUE),
            n = n()) %>%
  filter(var <= 1.7,
         n <= 1000,
         n >= 20) %>% 
  ggplot(aes(y = var, x = n)) +
  geom_point() +
  ggtitle("More than 10 samples") +
  theme_bw()
```

Differences in variance between sexes?
```{r}
dat %>% drop_na(hb, sex) %>% 
  group_by(sex, taxon) %>% 
  summarise(mean = mean(hb, na.rm = TRUE),
            var = var(hb, na.rm = TRUE),
            n = n()) %>% 
  filter(var <= 1.7,
         n <= 1000,
         n >= 20) %>% 
  ggplot(aes(y = var, x = n, color = sex)) +
  geom_point() +
  ggtitle("Differences in variance between sexes") +
  theme_bw()
```

Differences in bodysize variance between sexes? Try a model!
estimating the differences in hb length and hb length sigma between males and females
```{r}
d <- dat %>% select(taxon, sex, hb) %>% 
  drop_na(hb, sex)

m1 <- brm(bf(hb ~ 0 + sex,
             sigma ~ 0 + sex),
          prior = c(prior(normal(0, 1), class = "b"),
                    prior(normal(0, 1), class = "b", dpar = "sigma")),
          d,
          refresh = 0,
          chains = 2, cores = getOption("mc.cores", 4))

post <- posterior_samples(m1) %>% as_tibble() %>%
  select(starts_with("b_")) %>%
  mutate(dif_sex = b_sexf - b_sexm,
         #put sigma values on the right scale
         b_sigma_sexf = exp(b_sigma_sexf),
         b_sigma_sexm = exp(b_sigma_sexm),
         dif_sig = b_sigma_sexf - b_sigma_sexm) %>% 
  pivot_longer(cols = (1:6), names_to = "param", values_to = "val")

post %>% filter(param == c("dif_sex", "dif_sig")) %>% 
  ggplot(aes(x = val)) +
  geom_vline(xintercept = 0, alpha = 1/5) +
  stat_halfeye(.width = .89, alpha = 0.8,  normalize = "panels") +
  labs(x = "", y = "") +
  theme_bw() +
  facet_wrap(~ param, ncol = 1, labeller = as_labeller(c("dif_sex" = "Differences in Head-Body Length (F - M)", "dif_sig" = "Differences in Head-Body Length Variance"))) +
  ggtitle("Differences (Female - Male) in Body Size and Body Size Varaince (F - M")
  
```


# Intraspecific Variance Model

Havig some problems with the models converging. I think it has to do with the massive variance in the number of samples per species. I would guess that 21K samples would lead to BETTER convergence but I guess not. 


```{r}
d <- dat %>% select(taxon, sex, hb) %>% 
  drop_na(hb, sex)

m2 <- brm(bf(hb ~ 0 + sex + (1|taxon),
             sigma ~ 0 + sex),
          data = d,
          prior = c(prior(normal(-1, 1), class = "b"),
                    prior(normal(0, 1), class = "b", dpar = "sigma"),
                    prior(normal(0, 1), class = "sd")),
          refresh = 0,
          inits = 0,
          cores = getOption("mc.cores", 4),
          chains = 2,
          iter = 5000)

post2 <- posterior_samples(m2) %>% as_tibble() %>% 
  select(starts_with("b_")) %>% 
  mutate(dif_sex = b_sexf - b_sexm,
         #put sigma values on the right scale
         b_sigma_sexf = exp(b_sigma_sexf),
         b_sigma_sexm = exp(b_sigma_sexm),
         #calculate difference
         dif_sig = b_sigma_sexf - b_sigma_sexm) %>% 
  pivot_longer(cols = (1:6), names_to = "param", values_to = "val")

post2 %>% filter(param == c("dif_sex", "dif_sig")) %>% 
  ggplot(aes(x = val)) +
  geom_vline(xintercept = 0, alpha = 1/5) +
  stat_halfeye(.width = .89, alpha = 0.8, normalize = "panels") +
  labs(x = "", y = "") +
  theme_bw() +
  facet_wrap(~ param, ncol = 1, labeller = as_labeller(c("dif_sex" = "Differences in Head-Body Length (F - M)", "dif_sig" = "Differences in Head-Body Length Variance"))) +
  ggtitle("Differences (Female - Male) in Body Size and Body Size Varaince (F - M")
```
#### poor convergence

# Limit the number of samples per taxon

This leaves `n` samples per species (with higher counts than n). The others are left alone.
At `n=50` the convergence greatly improves. 
```{r}
n <- 50

d %>% group_by(taxon) %>% sample_n(if(n() < n) n() else n) -> d2

```


THe Trick here is modeling `~ 0 + sex + (0 + sex|taxon)` This estimates a different mean for each sex for each taxon rather than simply a different taxon mean. 


#### No Sigma Estimation Model
This works ok. 
```{r}
m2.1 <- brm(bf(hb ~ 0 + sex + (1|taxon)),
          data = d2,
          prior = c(prior(normal(0, 1), class = "b"),
                    #prior(normal(0, 1), class = "b", dpar = "sigma"),
                    prior(normal(0, 1), class = "sd")),
          refresh = 0,
          inits = 0,
          cores = getOption("mc.cores", 4),
          chains = 2,
          iter = 5000)

m2.1
```

#### Estimate Sigma
This particular model works well. A different sigma for each sex rather than a different sigma for each sex for each species.

```{r}
m2.2 <- brm(bf(hb ~ 0 + sex + (0 + sex|taxon),
               sigma ~ 0 + sex ),
          data = d2,
          prior = c(prior(normal(0, 1), class = "b"),
                    prior(normal(0, 1), class = "b", dpar = "sigma"),
                    prior(normal(0, 1), class = "sd")),
          refresh = 0,
          inits = 0,
          cores = getOption("mc.cores", 4),
          chains = 2,
          iter = 5000)

m2.2

post2.2 <- posterior_samples(m2.2) %>% as_tibble() %>% 
  select(starts_with("b_")) %>% 
  mutate(dif_sex = b_sexf - b_sexm,
         b_sigma_sexf = exp(b_sigma_sexf),
         b_sigma_sexm = exp(b_sigma_sexm),
         dif_sig = b_sigma_sexf - b_sigma_sexm
         ) %>% 
  #select(-"lp__") %>%
  pivot_longer(everything(), names_to = "param", values_to = "val")

post2.2 %>% filter(param == c("dif_sex", "dif_sig")) %>% 
  ggplot(aes(x = val)) +
  geom_vline(xintercept = 0, alpha = 1/5) +
  #geom_density(color = "firebrick", fill = "firebrick", names = c("CHanging", "names?")) +
  stat_halfeye(.width = .89, alpha = 0.8,  normalize = "panels") +
  labs(x = "", y = "") +
  theme_bw() +
  facet_wrap(~ param, ncol = 1, labeller = as_labeller(c("dif_sex" = "Differences in Head-Body Length (F - M)", "dif_sig" = "Differences in Head-Body Length Variance (F - M)"))) +
  ggtitle("Differences (Female - Male) in Body Size and Body Size Varaince\nMales are larger and varinace is higher in males")
```
