---
title: "R Notebook"
output: html_notebook
---
Scripts for testing differences in HB between sexes....

I wanted to keep this, but didn't want it in the `Initial_Data_Review.Rmd`. It needs work

Today I tried to look at differences in body-length between males and females while including taxon as a group-level effect. For some reason the grouping variable messes everything up. Probably due to the sheer number of some of the taxa in the set?

This is left today (3 November 2021) with 2 model versions, one without intraspecific variation, and one with. Interesting they give different answers! Without intra-variation, females are larger. With, males are larger. 

TO-DO:
-) Think about the phylogeny. Is this important? Use the Nations 2020 tree, even if it cuts the # of species by a lot. 
-) Still lousy convergence on the sex means. Is this just due to the range of body sizes? The priors? Not sure. 
-) In addition, why are the means for both males and females above zero??
-) Next I want to attach the locomotor mode info to the traits (the tree will help this!), and start working with intraspecific variance between locomotor modes. I think that arboreal taxa may have a lower intraspecific variance than "terrestrial" taxa. Based on Nations 2020, general were tightly conserved too. This is a great extension to that study. 


Differences in variance between sexes?
```{r}
dat %>% drop_na(hb, sex) %>% 
  group_by(sex, taxon) %>% 
  summarise(mean = mean(hb, na.rm = TRUE),
            var = var(hb, na.rm = TRUE),
            n = n()) %>% 
  filter(var <= 1.7,
         n <= 1000,
         n >= 20) %>% 
  ggplot(aes(y = var, x = n, color = sex)) +
  geom_point() +
  ggtitle("Differences in variance between sexes") +
  theme_bw()
```

Differences in bodysize variance between sexes? Try a model!
estimating the differences in hb length and hb length sigma between males and females
```{r}
d <- dat %>% select(taxon, sex, hb) %>% 
  drop_na(hb, sex) %>% 
  group_by(taxon) %>% 
  filter(n()>20) %>% 
  filter(n()<1000)


m1 <- brm(bf(hb ~ 0 + sex,
             sigma ~ 0 + sex),
          prior = c(prior(normal(0, 1), class = "b"),
                    prior(normal(0, 1), class = "b", dpar = "sigma")),
          d,
          refresh = 0,
          chains = 2, cores = getOption("mc.cores", 4))

post <- posterior_samples(m1) %>% as_tibble() %>%
  select(starts_with("b_")) %>%
  mutate(dif_sex = b_sexf - b_sexm,
         #put sigma values on the right scale
         b_sigma_sexf = exp(b_sigma_sexf),
         b_sigma_sexm = exp(b_sigma_sexm),
         dif_sig = b_sigma_sexf - b_sigma_sexm) %>% 
  pivot_longer(cols = (1:6), names_to = "param", values_to = "val")

post %>% filter(param == c("dif_sex", "dif_sig")) %>% 
  ggplot(aes(x = val)) +
  geom_vline(xintercept = 0, alpha = 1/5) +
  stat_halfeye(.width = .89, alpha = 0.8,  normalize = "panels") +
  labs(x = "", y = "") +
  theme_bw() +
  facet_wrap(~ param, ncol = 1, labeller = as_labeller(c("dif_sex" = "Differences in Head-Body Length (F - M)", "dif_sig" = "Differences in Head-Body Length Variance"))) +
  ggtitle("Differences (Female - Male) in Body Size and Body Size Varaince (F - M")
  
```


# Intraspecific Variance Model

Havig some problems with the models converging. I think it has to do with the massive variance in the number of samples per species. I would guess that 21K samples would lead to BETTER convergence but I guess not. 

Trying with student-t family and prior on sd. Robust regression is probably better for this type of analysis anyways!
```{r}
m2 <- brm(bf(hb ~ 0 + sex + (1|taxon),
             sigma ~ 0 + sex),
          family = skew_normal(),
          data = d,
          prior = c(prior(student_t(3, 0, 2.5), class = "b"),
                    prior(normal(0,1), class = "b", dpar = "sigma"),
                    prior(student_t(3, 0, 2.5), class = "sd")),
          refresh = 0,
          inits = 0,
          cores = getOption("mc.cores", 4),
          chains = 2,
          iter = 5000)

post2 <- posterior_samples(m2) %>% as_tibble() %>% 
  select(starts_with("b_")) %>% 
  mutate(dif_sex = b_sexf - b_sexm,
         #put sigma values on the right scale
         b_sigma_sexf = exp(b_sigma_sexf),
         b_sigma_sexm = exp(b_sigma_sexm),
         #calculate difference
         dif_sig = b_sigma_sexf - b_sigma_sexm) %>% 
  pivot_longer(cols = (1:6), names_to = "param", values_to = "val")

post2 %>% filter(param == c("dif_sex", "dif_sig")) %>% 
  ggplot(aes(x = val)) +
  geom_vline(xintercept = 0, alpha = 1/5) +
  stat_halfeye(.width = .89, alpha = 0.8, normalize = "panels") +
  labs(x = "", y = "") +
  theme_bw() +
  facet_wrap(~ param, ncol = 1, labeller = as_labeller(c("dif_sex" = "Differences in Head-Body Length (F - M)", "dif_sig" = "Differences in Head-Body Length Variance"))) +
  ggtitle("Differences (Female - Male) in Body Size and Body Size Varaince (F - M")
```
#### poor convergence
Too much variance on some of the taxa?
```{r}
d %>% group_by(taxon) %>% 
  summarise(var = var(hb)) %>% 
  arrange(desc(var))
```
Ok! The Mylomys dybowski and Dasymys incomtus have insanely large variance. Drop these and try again

I am also going to drop a few extreme hb values that seem like mistakes

```{r}
'%!in%' <- function(x,y)!('%in%'(x,y))
bad <- c("Mylomys_dybowskii", "Dasymys_incomtus")
d <- dat %>% select(taxon, sex, hb) %>% 
  drop_na(hb, sex) %>% 
  group_by(taxon) %>% 
  filter(n()>20) %>% 
  filter(n()<1000) %>% 
  filter(taxon %!in% bad) %>% 
  filter(hb < 4)
```

```{r}
m2.1 <- brm(bf(hb ~ 0 + sex + (1|taxon),
             sigma ~ 0 + sex),
          family = skew_normal(),
          data = d,
          prior = c(prior(student_t(3, 0, 2.5), class = "b"),
                    prior(normal(0,1), class = "b", dpar = "sigma"),
                    prior(student_t(3, 0, 2.5), class = "sd")),
          refresh = 0,
          inits = 0,
          cores = getOption("mc.cores", 4),
          chains = 2,
          iter = 5000)

```
#### Nested?
This is for a nested random effect, or a different estimate of sex values per taxon. 
```{r}
m2.2 <- brm(bf(hb ~ 0 + sex + (1|sex/taxon)
             #sigma ~ 0 + sex
             ),
          family = student(),
          data = d,
          prior = c(prior(normal(0,1), class = "b"),
                    prior(normal(0,1), class = "sigma"),
                    prior(normal(0,1), class = "sd")),
          refresh = 0,
          inits = 0,
          cores = getOption("mc.cores", 4),
          chains = 2,
          iter = 2500)
plot(m2.2, N=3, ask=F)
```


```{r}
m2.3 <- brm(bf(hb ~ 0 + sex + (1|gr(taxon, by = sex))
             #sigma ~ 0 + sex
             ),
          family = student(),
          data = d,
          prior = c(prior(normal(0,1), class = "b"),
                    prior(normal(0,1), class = "sigma"),
                    prior(normal(0,1), class = "sd")),
          refresh = 0,
          inits = 0,
          cores = getOption("mc.cores", 4),
          chains = 2,
          iter = 2500)
plot(m2.3, N=3, ask=F)
```



```{r}
post2 <- posterior_samples(m2.2) %>% as_tibble() %>% 
  select(starts_with("b_")) %>% 
  mutate(dif_sex = b_sexf - b_sexm#,
         #put sigma values on the right scale
         #b_sigma_sexf = exp(b_sigma_sexf),
         #b_sigma_sexm = exp(b_sigma_sexm),
         #calculate difference
         #dif_sig = b_sigma_sexf - b_sigma_sexm
         ) %>% 
  pivot_longer(cols = (1:6), names_to = "param", values_to = "val")

post2 %>% filter(param == c("dif_sex")) %>% 
  ggplot(aes(x = val)) +
  geom_vline(xintercept = 0, alpha = 1/5) +
  stat_halfeye(.width = .89, alpha = 0.8, normalize = "panels") +
  labs(x = "", y = "") +
  theme_bw() +
  #facet_wrap(~ param, ncol = 1, labeller = as_labeller(c("dif_sex" = "Differences in Head-Body Length (F - M)", "dif_sig" = "Differences in Head-Body Length Variance"))) +
  ggtitle("Differences (Female - Male) in Body Size and Body Size Varaince (F - M")
```


```{r}
m3 <- brm(bf(hb ~ 0 + sex + (0 + sex|taxon),
               sigma ~ 0 + sex ),
          data = d,
          family = student(),
          prior = c(prior(normal(0, 1), class = "b"),
                    #prior(normal(0, 1), class = "b", dpar = "sigma"),
                    prior(normal(0, 1), class = "sd")),
          refresh = 0,
          inits = 0,
          cores = getOption("mc.cores", 4),
          chains = 2,
          iter = 5000)

m3
```

```{r}
post3 <- posterior_samples(m3) %>% as_tibble() %>% 
  select(starts_with("b_")) %>% 
  mutate(dif_sex = b_sexf - b_sexm,
         #put sigma values on the right scale
         b_sigma_sexf = exp(b_sigma_sexf),
         b_sigma_sexm = exp(b_sigma_sexm),
         #calculate difference
         dif_sig = b_sigma_sexf - b_sigma_sexm) %>% 
  pivot_longer(cols = (1:6), names_to = "param", values_to = "val")

post3 %>% filter(param == c("dif_sex", "dif_sig")) %>% 
  ggplot(aes(x = val)) +
  geom_vline(xintercept = 0, alpha = 1/5) +
  stat_halfeye(.width = .89, alpha = 0.8, normalize = "panels") +
  labs(x = "", y = "") +
  theme_bw() +
  facet_wrap(~ param, ncol = 1, labeller = as_labeller(c("dif_sex" = "Differences in Head-Body Length (F - M)", "dif_sig" = "Differences in Head-Body Length Variance (F - M)"))) +
  ggtitle("Differences (Female - Male) in Body Size and Body Size Varaince (F - M)")
```

# Limit the number of samples per taxon

This leaves `n` samples per species (with higher counts than n). The others are left alone.
At `n=50` the convergence greatly improves. 
```{r}
n <- 200

d %>% group_by(taxon) %>% sample_n(if(n() < n) n() else n) -> d2

```

```{r}
m4 <- brm(bf(hb ~ 0 + sex + (0 + sex|taxon),
               sigma ~ 0 + sex ),
          data = d2,
          family = student(),
          prior = c(prior(normal(0, 1), class = "b"),
                    #prior(normal(0, 1), class = "b", dpar = "sigma"),
                    prior(normal(0, 1), class = "sd")),
          refresh = 0,
          inits = 0,
          cores = getOption("mc.cores", 4),
          chains = 2,
          iter = 5000)

m4
```
```{r}
post4 <- posterior_samples(m4) %>% as_tibble() %>% 
  select(starts_with("b_")) %>% 
  mutate(dif_sex = b_sexf - b_sexm,
         #put sigma values on the right scale
         b_sigma_sexf = exp(b_sigma_sexf),
         b_sigma_sexm = exp(b_sigma_sexm),
         #calculate difference
         dif_sig = b_sigma_sexf - b_sigma_sexm) %>% 
  pivot_longer(cols = (1:6), names_to = "param", values_to = "val")

post4 %>% filter(param == c("dif_sex", "dif_sig")) %>% 
  ggplot(aes(x = val)) +
  geom_vline(xintercept = 0, alpha = 1/5) +
  stat_halfeye(.width = .89, alpha = 0.8, normalize = "panels") +
  labs(x = "", y = "") +
  theme_bw() +
  facet_wrap(~ param, ncol = 1, labeller = as_labeller(c("dif_sex" = "Differences in Head-Body Length (F - M)", "dif_sig" = "Differences in Head-Body Length Variance (F - M)"))) +
  ggtitle("Differences in Body Size and Varaince (F - M)\n200 samples per species")
```

# Species Means
run the same test but with a mean for each sex for each species
```{r}
d_mean <- d %>% 
  group_by(sex, taxon) %>% 
  summarise(mean_hb = mean(hb, na.rm = TRUE)) %>% 
  ungroup()
```

```{r}
m5 <- brm(bf(mean_hb ~ 0 + sex ,
               sigma ~ 0 + sex ),
          data = d_mean,
          family = student(),
          prior = c(prior(normal(0, 1), class = "b"),
                    prior(normal(0, 1), class = "b", dpar = "sigma")),
          refresh = 0,
          inits = 0,
          cores = getOption("mc.cores", 4),
          chains = 2,
          iter = 5000)
```

```{r}
post5 <- posterior_samples(m5) %>% as_tibble() %>% 
  select(starts_with("b_")) %>% 
  mutate(dif_sex = b_sexf - b_sexm,
         #put sigma values on the right scale
         b_sigma_sexf = exp(b_sigma_sexf),
         b_sigma_sexm = exp(b_sigma_sexm),
         #calculate difference
         dif_sig = b_sigma_sexf - b_sigma_sexm) %>% 
  pivot_longer(cols = (1:6), names_to = "param", values_to = "val")

post5 %>% filter(param == c("dif_sex", "dif_sig")) %>% 
  ggplot(aes(x = val)) +
  geom_vline(xintercept = 0, alpha = 1/5) +
  stat_halfeye(.width = .89, alpha = 0.8, normalize = "panels") +
  labs(x = "", y = "") +
  theme_bw() +
  facet_wrap(~ param, ncol = 1, labeller = as_labeller(c("dif_sex" = "Differences in Head-Body Length (F - M)", "dif_sig" = "Differences in Head-Body Length Variance (F - M)"))) +
  ggtitle("Differences in Body Size and Varaince (F - M)\nSpecies Means")
```


THe Trick here is modeling `~ 0 + sex + (0 + sex|taxon)` This estimates a different mean for each sex for each taxon rather than simply a different taxon mean. 


#### No Sigma Estimation Model
This works ok. 
```{r}
m2.1 <- brm(bf(hb ~ 0 + sex + (1|taxon)),
          data = d2,
          family = student(),
          prior = c(prior(normal(0, 1), class = "b"),
                    #prior(normal(0, 1), class = "b", dpar = "sigma"),
                    prior(normal(0, 1), class = "sd")),
          refresh = 0,
          inits = 0,
          cores = getOption("mc.cores", 4),
          chains = 2,
          iter = 5000)

m2.1
```

#### Estimate Sigma
This particular model works well. A different sigma for each sex rather than a different sigma for each sex for each species.

```{r}
m2.2 <- brm(bf(hb ~ 0 + sex + (0 + sex|taxon),
               sigma ~ 0 + sex ),
          data = d2,
          prior = c(prior(normal(0, 1), class = "b"),
                    prior(normal(0, 1), class = "b", dpar = "sigma"),
                    prior(normal(0, 1), class = "sd")),
          refresh = 0,
          inits = 0,
          cores = getOption("mc.cores", 4),
          chains = 2,
          iter = 5000)

m2.2

post2.2 <- posterior_samples(m2.2) %>% as_tibble() %>% 
  select(starts_with("b_")) %>% 
  mutate(dif_sex = b_sexf - b_sexm,
         b_sigma_sexf = exp(b_sigma_sexf),
         b_sigma_sexm = exp(b_sigma_sexm),
         dif_sig = b_sigma_sexf - b_sigma_sexm
         ) %>% 
  #select(-"lp__") %>%
  pivot_longer(everything(), names_to = "param", values_to = "val")

post2.2 %>% filter(param == c("dif_sex", "dif_sig")) %>% 
  ggplot(aes(x = val)) +
  geom_vline(xintercept = 0, alpha = 1/5) +
  #geom_density(color = "firebrick", fill = "firebrick", names = c("CHanging", "names?")) +
  stat_halfeye(.width = .89, alpha = 0.8,  normalize = "panels") +
  labs(x = "", y = "") +
  theme_bw() +
  facet_wrap(~ param, ncol = 1, labeller = as_labeller(c("dif_sex" = "Differences in Head-Body Length (F - M)", "dif_sig" = "Differences in Head-Body Length Variance (F - M)"))) +
  ggtitle("Differences (Female - Male) in Body Size and Body Size Varaince\nMales are larger and varinace is higher in males")
```
